NumPy---Array Programming in Python

### 2020-02-14 body.tex
Minor edits

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its significant application, such as the
\code{eht-imaging} library, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the primary}
software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

In recent years, there has been a rapid growth in the development of
specialized hardware, including graphics processing
units (GPUs), tensor processing units (TPUs), and field-programmable
gate arrays (FPGAs).  Data sets now routinely exceed the memory capacity of a single machine and may be stored on multiple machines or in
the cloud.  Due to its in-memory data model, NumPy is currently unable to
utilize such specialized hardware and storage directly.  However, both
distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
to the \emph{paradigm} of array programming: a gap, therefore, existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. For example, each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and PyData/Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for specialized arrays to be
passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
NumPy, in turn, dispatches operations to the originating library, as required.
Over four hundred of the most popular
NumPy functions are supported.  The protocols are implemented by at
least five libraries, including Dask, CuPy, xarray, and PyData/Sparse.
Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask.
The protocols \textit{compose} well, and so it is possible to redeploy NumPy
code onto multi-GPU systems with minimal changes\cite{entschev2019}.
Working with the community, we have identified several issues that
prevent further adoption, and are refining interfaces to reduce
impedance mismatch.

@@

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the broader scientific Python ecosystem and
heralded the current era of widespread use of Python for scientific
computing. % This and the "What began as" sentence below are more or less interchangeable.

%What began as an attempt to add an array object to Python became
%the foundation of a vibrant ecosystem of tools.
%Now, a large amount of scientific work depends on NumPy being correct,
%fast, and stable.
%%
%%
%% v TO ADD v
%%
%% Over the next decade, the computational landscape will change
%% drastically.
%% - AI / data science mature
%% - hardware / Moore's law, specialized devices
%% - more users (diversity + numbers)
%% - scale / type of data; scientific data, lightsheet microscopes, LSST
%% - new ideas come from other languages like Julia, Rust, ...; JIT improvements
%% NumPy is poised to tackle these challenges, and work continues.
%% - Major refactoring of code
%% - Governance
%% - Process (NEP, roadmap, how we handle issues on GH, etc.)
%% - Protocols: will be extending that work
%% - Funding (some recent evidence that that is improving → grants, or
%% Moore/Sloan/CZI) CZI EOSS
%% - Redesigning website and documentation to be more friendly, easy to
%% navigate
%% - Next generation of developers (grad students) to sign on
%%

@@

>>>>

The recent rapid growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its significant application, such as the
\code{eht-imaging} library, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the primary}
software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

However, scientific data sets now routinely exceed the memory capacity of a single machine and may
be stored on multiple machines or in the cloud.
In addition, the recent need to accelerate deep learning and artificial intelligence applications
has led to the emergence of specialized accelerator hardware,
including graphics processing units (GPUs), tensor processing units (TPUs),
and field-programmable gate arrays (FPGAs).
Due to its in-memory data model, NumPy is currently unable to
utilize such storage and specialized hardware directly.  However, both
distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
to the \emph{paradigm} of array programming: a gap, therefore, existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

The community's efforts to fill this gap led to a
proliferation of new array implementations. For example, each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and PyData/Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for specialized arrays to be
passed to NumPy functions (Fig.~\ref{fig:array-protocol}).
NumPy, in turn, dispatches operations to the originating library, as required.
Over four hundred of the most popular
NumPy functions are supported.
The protocols are implemented by widely used libraries such as Dask, CuPy,
xarray, and PyData/Sparse.
Thanks to these developments, users can now, for example, scale
their computation from a single machine to distributed systems using Dask.
The protocols also \textit{compose} well, allowing users to redeploy NumPy
code at scale on distributed, multi-GPU systems via, for instance, CuPy arrays embedded in Dask
arrays. Using NumPy's high-level API, users can leverage highly parallel code
execution on multiple systems with millions of cores, all with minimal code
changes \cite{entschev2019}.

These array protocols are now a key feature of NumPy, and are expected to only
increase in importance.  As with the rest of NumPy, we iteratively refine and
add protocol designs to improve utility and simplify adoption.

@@

%NumPy has subsequently seen prolific adoption, with millions of users,
%first from research and later also industry.
%It spurred the development of the broader scientific Python ecosystem and
%heralded the current era of widespread use of Python for scientific
%computing. % This and the "What began as" sentence below are more or less interchangeable.

What began as an attempt to add an array object to Python became
the foundation of a vibrant ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
It is no longer a small community project, but is core scientific
infrastructure.
The developer culture has changed as well.
- Governance
- Process (NEP, roadmap, how we handle issues on GH, etc.)
- still primarily draw from graduate students and researchers
- Funding (some recent evidence that that is improving  grants, or
Moore/Sloan/CZI) CZI EOSS
- Major refactoring of code
- Redesigning website and documentation to be more friendly, easy to
navigate

@@

- Protocols: will be extending that work

Over the next decade, we will face new challenges.
- AI / data science mature
- hardware / Moore's law, specialized devices
- more users (diversity + numbers)
- scale / type of data; scientific data, lightsheet microscopes, LSST
- new ideas come from other languages like Julia, Rust, ...; JIT improvements

NumPy is poised to tackle these challenges, and florish in the next decade.
Need
- Next generation of developers (grad students) to sign on
- Funders and industry continue funding this project as well as increasing support for the ecosystem through new funding mechanisms.
